Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-23 icfp.tex
done 3.2

Linear Core ($\lambda^\pi_\Delta$) is a linear lambda calculus akin to Linear
Haskell's $\lambda^q_\to$ in that both have multiplicity polymorphism,
(recursive) let bindings, case expressions, and algebraic data types.
$\lambda^\pi_\Delta$ diverges from $\lambda^q_\to$ primarily when typing lets,
case expressions, and alternatives (in its purpose to type semantic linearity).
%, and secondarily in only treating multiplicity polymorphism superficially.
%
% \todo{And our treatment of multiplicity polymorphism completely ignores
% algebraic treatment of multiplicities with semiring operations!!! Would it be
% sufficient to add a rule for application of variable multiplicity functions?}
%
%
The core rules of the calculus for, multiplicity and term,
abstraction and application are similar to that of $\lambda^q_\to$ and
mostly standard. 
We note that we handle multiplicity polymorphism differently from
Linear Haskell in ignoring the multiplicity semiring and, instead,
conservatively treating all multiplicity polymorphic functions as linear.
%
The full type system is given in Figure~\ref{fig:linear-core-typing-rules},
with auxiliary judgements given in
Figure~\ref{fig:linear-core-other-judgements}.

@@

We start with the main typing judgement.
As is common for linear type
systems, we use two typing environments, an \emph{unrestricted} $\G$ and
a \emph{linear} $\D$ environment.
Thus, the main judgement is written $\G;\D \vdash e :
\s$,
denoting that expression $e$ has type $\s$ under
the unrestricted environment $\G$ and linear environment $\D$.

%
Variables in $\G$ can be freely discarded (\emph{weakened}) and
duplicated (\emph{contracted}), while resources in $\D$ must be used
exactly once (hence can't be discarded nor duplicated). Despite not
having explicit weakening and contraction rules in our system, they
are available as admissible rules for $\G$ (but not for $\D$), since,
equivalently \cite{91621fae-5e53-3497-8291-32b2fab5a743}, resources
from $\G$ are duplicated for sub-derivations and may unrestrictedly
exist in the variable rules.
% TODO: Likely rewrite that
%

@@

Linear functions are introduced via the function type ($\s \to_\pi \vp$) with
$\pi = 1$, i.e. a function of type $\s \to_1 \vp$ (or, equivalently, $\s \lolli \vp$)
introduces a linear resource of type $\s$ in the linear environment $\D$ to then type an expression of type $\varphi$.
%
Unrestricted functions are introduced via the function type ($\s \to_\pi \vp$) with $\pi =
\omega$, and the $\lambda$-bound variable is introduced in $\G$ (rules
$\lambda I_1$ and $\Lambda I$). Function application is standard, with
linear function application splitting the linear context in two
disjoint parts, one used to type the argument and the other for the
function.
Arguments of unrestricted functions must also be unrestricted, i.e. no
linear variables can be used to type them. 
% \[
% \begin{array}{cc}
% \TypeLamIntroL & \TypeLamIntroW
% \end{array}
% \]
% A linear function application is well-typed if there exists a disjoint split of the
% linear resources into $\D,\D'$ s.t. the function and argument, each under a
% distinct split, are both well-typed and the argument type matches the
% function's expected argument type. Conversely, unrestricted resources are
% duplicated and available whole to both sub-derivations.
% \[
% \TypeLamElimL \quad \TypeLamElimW
% \]
% An unrestricted function, unlike a linear one, consumes its argument
% unrestrictedly (zero or more times). Therefore, in an unrestricted function
% application, allowing any linear resources to occur in the argument expression
% entails consuming those resources unrestrictedly, since the
% variable binding the argument expression could be discarded or used more than
% once in the function body. Thus, argument expressions to unrestricted functions must also
% be unrestricted, i.e. no linear variables can be used to type them.
% %
%
Typing linear and unrestricted function applications separately is less general
than typing applications of functions of any multiplicity $\pi$ by scaling (per the multiplicity
semiring) the multiplicities of the resources used to type the argument by $\pi$, however, our objective of typing semantic
linearity does not benefit much from doing so and therefore we opt for a
simpler design. 
% \todo{We might only need multiplicities for the case-of-case definition as it
% exists in ghc?, even then couldn't we do semiring scaling without variables?
% Also, dislike sentence}

Multiplicity abstractions ($\Lambda p.~e$) introduce a multiplicity variable
$p$ (in the unrestricted context), and construct expressions of type
$\forall p.~\s$, i.e. a type universally quantified over a multiplicity
variable $p$. Again, conservatively, in the body of the abstraction, function
types annotated with a $p$ variable and datatype fields with multiplicity $p$
are typed as thought they are linear functions and linear fields, because $p$
can be instantiated at both $\omega$ and $1$.
\[
\TypeMultLamIntro \quad \TypeMultLamElim
\]
A multiplicity application instantiates a multiplicity-polymorphic type
$\forall p.~\s$ at a particular (argument) multiplicity $\pi$, resulting in an
expression of type $\s$ where occurrences of $p$ are substituted by $\pi$, i.e.
$\s[\pi/p]$.
The rule additionally requires that $\pi$ be \emph{well-formed} in order
for the expression to be well-typed, using the judgement $\G \vdash_{mult}
\pi$, where well-formedness is given by $\pi$ either being $1$, $\omega$, or an
in-scope  multiplicity variable in $\G$.
% \[
% \TypeWellFormedMult
% \]

>>>>

Linear Core ($\lambda^\pi_\Delta$) is a linear lambda calculus akin to Linear
Haskell's $\lambda^q_\to$ in that both have multiplicity polymorphism,
(recursive) let bindings, case expressions, and algebraic data types.
$\lambda^\pi_\Delta$ diverges from $\lambda^q_\to$ primarily when typing lets,
case expressions, and alternatives (in its purpose to type semantic linearity).
%, and secondarily in only treating multiplicity polymorphism superficially.
%
% \todo{And our treatment of multiplicity polymorphism completely ignores
% algebraic treatment of multiplicities with semiring operations!!! Would it be
% sufficient to add a rule for application of variable multiplicity functions?}
%
%
The core rules of the calculus for abstraction and application are similar to
that of $\lambda^q_\to$ and mostly standard. 
We note that we handle multiplicity polymorphism differently from
Linear Haskell in ignoring the multiplicity semiring and, instead,
conservatively treating all multiplicity polymorphic functions as linear.
%
The full type system is given in Figure~\ref{fig:linear-core-typing-rules},
with auxiliary judgements given in
Figure~\ref{fig:linear-core-other-judgements}.

@@

%We start with the main typing judgement.
%As is common for linear type systems, we use two typing environments, an
%unrestricted $\G$ and a linear $\D$ environment.
The main judgement is written $\G;\D \vdash e : \s$ to denote that expression
$e$ has type $\s$ under the unrestricted environment $\G$ and linear
environment $\D$.
%
%
Variables in $\G$ can be freely discarded (\emph{weakened}) and duplicated
(\emph{contracted}), while resources in $\D$ must be used exactly once. Despite
not having explicit weakening and contraction rules in our system, they are
available as admissible rules for $\G$ (but not for $\D$), since, equivalently
\cite{91621fae-5e53-3497-8291-32b2fab5a743}, resources from $\G$ are duplicated
for sub-derivations and may unrestrictedly exist in the variable rules.
% TODO: Likely rewrite that?
%

@@

Linear functions are introduced via the function type ($\s \to_\pi \vp$) with
$\pi = 1$, i.e. a function of type $\s \to_1 \vp$ (or, equivalently, $\s \lolli \vp$)
introduces a linear resource of type $\s$ in the linear environment $\D$ to then type an expression of type $\varphi$.
%
Unrestricted functions are introduced via the function type ($\s \to_\pi \vp$) with $\pi =
\omega$, and the $\lambda$-bound variable is introduced in $\G$ (rules
$\lambda I_1$ and $\Lambda I$). Function application is standard, with
linear function application splitting the linear context in two
disjoint parts, one used to type the argument and the other the
function.
Arguments of unrestricted functions must also be unrestricted, i.e. no
linear variables can be used to type them. 
% \[
% \begin{array}{cc}
% \TypeLamIntroL & \TypeLamIntroW
% \end{array}
% \]
% A linear function application is well-typed if there exists a disjoint split of the
% linear resources into $\D,\D'$ s.t. the function and argument, each under a
% distinct split, are both well-typed and the argument type matches the
% function's expected argument type. Conversely, unrestricted resources are
% duplicated and available whole to both sub-derivations.
% \[
% \TypeLamElimL \quad \TypeLamElimW
% \]
% An unrestricted function, unlike a linear one, consumes its argument
% unrestrictedly (zero or more times). Therefore, in an unrestricted function
% application, allowing any linear resources to occur in the argument expression
% entails consuming those resources unrestrictedly, since the
% variable binding the argument expression could be discarded or used more than
% once in the function body. Thus, argument expressions to unrestricted functions must also
% be unrestricted, i.e. no linear variables can be used to type them.
% %
%
Typing linear and unrestricted function applications separately is less general
than typing applications of functions of any multiplicity $\pi$ by scaling (per the multiplicity
semiring) the multiplicities of the resources used to type the argument by $\pi$, however, our objective of typing semantic
linearity does not benefit much from doing so and therefore we opt for a
simpler design. 
% \todo{We might only need multiplicities for the case-of-case definition as it
% exists in ghc?, even then couldn't we do semiring scaling without variables?
% Also, dislike sentence}

Multiplicity abstractions ($\Lambda I$) introduce a multiplicity variable $p$,
and construct expressions of type $\forall p.~\s$, i.e. a type universally
quantified over a multiplicity variable $p$. Again, conservatively, in the body
of the abstraction, function types annotated with a $p$ variable and datatype
fields with multiplicity $p$ are typed as thought they are linear functions and
linear fields, because $p$ can be instantiated at both $\omega$ and $1$.
%\[
%%\TypeMultLamIntro \quad \TypeMultLamElim
%\]
A multiplicity application ($\Lambda E$) instantiates a multiplicity-polymorphic type
$\forall p.~\s$ at a particular (argument) multiplicity $\pi$, resulting in an
expression of type $\s$ where occurrences of $p$ are substituted by $\pi$, i.e.
$\s[\pi/p]$.
The rule additionally requires that $\pi$ be \emph{well-formed} in order
for the expression to be well-typed, using the judgement $\G \vdash_{mult}
\pi$, where well-formedness is given by $\pi$ either being $1$, $\omega$, or an
in-scope  multiplicity variable in $\G$.
% \[
% \TypeWellFormedMult
% \]
